RCloud: Integrating Exploratory Visualization, Analysis and Deployment

### 2014-03-28 conclusion.tex
yes

We presented the ideas that motivated the design of RCloud,
a prototype visual analytics environment to support
collaboration, searching, sharing, and publishing of
experiments, similar to DevOps development of web services.
Experience with the prototype provided evidence that
data science teams and the organizations in which they work
can benefit from having such features.

>>>>

We described the motivation for RCloud, a prototype
visual analytics environment that supports collaboration,
searching, sharing, and publishing of experiments.
This assists software development and release using
practices similar to DevOps in web services.
Experience with the Rcloud prototype provided evidence that
data science teams and the organizations in which they work
benefit from having such capabilities.

### 2014-03-28 discussion.tex
yes

Experience with deploying RCloud in a community of data analysts
has provided some insight about how the proposed requirements
were met, and to what extent they were appropriate.
This experience may be informative to researchers and designers
of future systems supporting collaborative visual analytics.

Returning to the design considerations defined by Heer and Agrawala,
our experience underscores the importance of many of their proposals,
and indicates areas for further exploration.

{\bf Shared artifacts and artifact histories} are a central feature
of RCloud. The prototype aimed at making it convenient to collaborate
on experimental workbooks and the analyses that result. We observed
that analysts readily shared ideas through RCloud. They used RCloud
to discuss data analysis, programming problems, and to publish results
by sharing URLs instead of instead of embedding screenshots in
PowerPoint.\todo{Talk about artifact histories?}

@@

{\bf Discussion.} Annotation is supported in our human interface,
and was another central goal of our prototype. \todo{Elaborate.}
An interesting question is, to what extent should users of production
notebooks also able to use annotation and similar tools without coding
and without being exposed to the analyst's view?\todo{Is this a gap in
our design?}

{\bf Content export} is not a capability we worked to support,
but the R environment already has many packages for this.
Our goal was to devise ways of working where explicit export
is less necessary. However it is desirable to integrate results
through web protocols such as JSON. We did assume that RCloud
notebooks need to play well in the ecosystem of other tools,
and this worked out as intended, though we learned it also creates
a temptation to absorb features almost without limit. (Should
RCloud support templates for personal web pages and graphical
editing? Should it incorporate an email client and emacs?
Should it be able to push results to Facebook?)

{\bf Social-psychological incentives} and {\bf voting and ranking}
are supported through starring. This mechanism was employed often
by our test audience. An obvious next step would be to enhance
recommendations via relationships discovered by static and
dynamic code analysis. This should be done both within and across
collections of scripts (the latter being similar to Vistrails'
enhanced recommendations by clustering multiple workflows).
It should be valuable to data analysts to know which packages are
frequently used together, or appear with a given record type
or data feed. In general, though, trivial or passive mechanisms
to collect data for recommendations are essential.

@@

Because we developed the ideas and the prototype together,
we did not foresee some of the requirements that emerged
after our audience started working with the prototype.

@@

A disadvantage is that some important aspects of the environment
are more difficult to manage in RCloud than in conventional systems.
Not only does it lack an explicit separation of the development and
deployment environments, but every version of every prototype is
shared by default. While this encourages collaboration on work
in progress, errors such as misconfiguration, mismatches between
versions of packages, and ordinary programming errors are also
easily shared and can affect production websites unintentionally.
This cannot be avoided entirely, but the power of sharing is worth
this price (as proven true with distributed systems in general).
Still, this problem needs more attention if the environment is
going to be scaled up to a wider number of users with fewer
informal channels to coordinate work. \todo{need a transition}
Versioning occurs in several places: in scripts, in the R environment
such as the installed R libraries and packages, and in the external
environment such as the operating system and even protocols spoken by
remote services. At one end of the scale, we have full control over the
versioning of scripts via git, though there is still a need for conventions
for example to name stable or working versions of scripts or versions known
to work together. The R environment is under the control of its package
manager. It is at least possible for RCloud to access information about
package versions and configurations, but support for checking compatibility
and maintaining multiple versions in the same environment is not strong.
At the other end of the scale, there is not much reason to expect version
control for the external environment. Most programmers rely on clean living
and a conservative approach to upgrades.

On top of this, RCloud allows analysts to fork experiments to try
new ideas quickly. But we have not done much so far to address the
problem of having a lot of bits and pieces of code and data lying
around, though at least we have made it easier to get at them and
to apply algorithms and metadata to organize them.

>>>>

Experience with deploying RCloud in a community of data analysts
(hackers in the parlance of Heer and Agrawala)
has provided some insight about how the proposed requirements
were met, and to what extent they were appropriate.
This experience may be informative to researchers and designers
of future systems supporting collaborative visual analytics.

Returning to the design considerations defined by Heer and Agrawala,
our experience underscores the importance of their proposals,
and indicates areas for further exploration, especially in the
following areas.

{\bf Shared artifacts and artifact histories} are a central feature
of RCloud. The prototype aimed at making it convenient to collaborate
on experimental workbooks and the resulting analyses. We observed
that hackers readily shared ideas through RCloud. They used RCloud
to discuss data analysis, programming problems, and to publish results
by sharing URLs instead of instead of embedding screenshots in
PowerPoint. Artifact histories are provided through git. It is fair
to criticize that it is not human-friendly, but we assume RCloud
hackers are already familiar with it.

{\bf Modularity and granularity.} The ability to segment work into
independent units (modularity) is a key to working in teams. It is
further an advantage if the units may be kept small, and team members
can realize benefits that are at least proportional to their work
on these units (granularity). RCloud's notebook and versioning
capabilities allow work to be divided into units as fine as
the underlying language allows, and versioning encourages making
incremental changes at low cost and without disrupting the work
of others.

@@

{\bf Discussion.} Annotation and commenting was another central goal
of our prototype, and is provided in our human interface.
An interesting question is, to what extent should application users
be able to make annotations and operate similar tools in published
notebooks without coding and being exposed to the hackers's view?
Our prototype has not addressed this problem yet.

{\bf Content export} is not a capability we worked to support,
but R already has many packages for this. Our goal was to devise
ways of working where explicit export is less necessary.
However it is desirable to integrate results
through web protocols such as JSON. We did assume that RCloud
notebooks need to play well in the ecosystem of other tools,
and this worked out as intended, though we learned it also creates
a temptation to absorb features almost without limit. (Should
RCloud support templates for personal web pages and graphical
editing? Should it incorporate an email client and emacs?
Should it be able to push results to Facebook?)

{\bf Social-psychological incentives} and {\bf voting and ranking}
are supported through starring. This mechanism was employed often
by our test audience. An obvious next step would be to enhance
recommendations via relationships discovered by static and
dynamic code analysis. This should be done both within and across
collections of scripts (the latter being similar to Vistrails'
enhanced recommendations by clustering multiple workflows).
It should be valuable to hackers to know which packages are
frequently used together, or appear with a given record type
or data feed. In general, though, trivial or passive mechanisms
to collect data for recommendations are essential.

@@

Because we developed the ideas and the prototype together,
we did not foresee some of the requirements that emerged
after people started working with the prototype.

@@

A disadvantage is that some important aspects of the environment
are more difficult to manage in RCloud than in conventional systems.
Not only does it lack an explicit separation of the development and
deployment environments, but every version of every prototype is
shared by default. While this encourages collaboration on work
in progress, errors such as misconfiguration, mismatches between
versions of packages, and ordinary programming errors are also
easily shared and can affect production websites unintentionally.
This cannot be avoided entirely, but the power of sharing is worth
this price (as proven true with distributed systems in general).
Still, this problem needs more attention if the environment is
going to be scaled up to a wider number of users with fewer
informal channels to coordinate work. Effective sharing in RCloud
will require control over versioning.
Versioning must be managed in several places: in scripts, in the R environment
such as the installed R libraries and packages, and in the external
environment such as the operating system and even protocols spoken by
remote services. At one end of the scale, we have full control over the
versioning of scripts via git, though there is still a need for conventions
for example to name stable or working versions of scripts or versions known
to work together. The R environment is under the control of its package
manager. It is at least possible for RCloud to access information about
package versions and configurations, but support for checking compatibility
and maintaining multiple versions in the same environment is not strong.
At the other end of the scale, there is not much reason to expect version
control for the external environment. Most programmers rely on clean living
and a conservative approach to upgrades.

On top of this, RCloud allows hackers to fork experiments to try
new ideas quickly. So far we have not done much more to address the
problem of having a lot of bits and pieces of code and data lying
around, though at least we have made it easier to get at them and
to apply algorithms and metadata to organize them.

### 2014-03-28 introduction.tex
yes

Second, EDA environments often assume a single-developer perspective,
while data scientist teams could greatly benefit from convenient searching
and sharing of scripts, data sets, data feeds, experiments, annotations,
and automated analyses and recommendations. These capabilities have had
tremendous impact on information retrieval in general but are well
beyond what existing visual analytics environments provide, or what
is found in today's repositories and version control systems.

@@

\begin{itemize}
\item Constant technology transfer. Experiments in notebooks developed
by analysts need to become accessible in the production environment,
making it easy to push new features, prototypes, etc.
\item Dashboarding. To support the above, there should be a mode
in which non-experts can operate the experiments without coding.
This may require some accomodation in code written by analysts,
but our goal is to make that as simple as we can.
\item Version control as a primary concept in the repository. There
needs to be a way to control the visibility of prototypes as they
are developed and tested.
\item Coexistence in the current data ecosystem. The isolation of
today's exploratory visualization and data analysis environments is
a real obstacle to wider adoption of novel data analysis methods.
\item Modern interactive technologies. Current isolation notwithstanding,
there is demand for interactive visualization tools. In some respects,
the prevalence of data journalism has raised expectations very high.
The main challenge is how to integrate these technologies so data
scientists and statisticians can program in them, \emph{and} business-facing
workers can operate them.
%% Scalable technologies like crossfilter exist. Want to support those in the context of an R environment. Need to support full 2-way communication between computing in the cloud and interaction on the web.
%% \item Other requirements e.g. Kandel's user study. Not our contribution but things we must address, mentioned in related work section. Heer/Agrawala design considerations for collaborative visual analytics.
\end{itemize}

% What is the scope of our solution?
Our investigation focuses on the R language \cite{RCoreTeam:2013:R}
for statistical computing and graphics. R is one of the most popular
data analysis systems and is well-supported by a complex ecosystem
of compatible packages, software tools and other resources.
%
Of importance to us, there are many existing R packages, including
packages for high performance computing in the cloud, domain specific
models,
and for generating web graphics. % no, there aren't. we wrote those!
% well, right about that

% We want our solution to play well with the rest of the ecosystem

>>>>

Second, EDA environments often assume a single-developer perspective,
while data scientist teams could greatly benefit from convenient searching
and sharing of scripts, data sets, data feeds, experiments, annotations,
and automated analyses and recommendations. These capabilities have had
tremendous impact on information retrieval in general but are well
beyond what existing visual analytics environments provide, or what
is found in current repositories and version control systems.

@@

\begin{itemize}
\item Constant technology transfer. Experiments in notebooks developed
by analysts need to become accessible in the production environment,
making it easy to push new features, prototypes, etc.
\item Dashboarding. To support the above, there should be a mode
in which non-experts can operate the experiments without coding.
This may require some accomodation in code written by analysts,
but our goal is to make that as simple as we can.
\item Version control as a primary concept in the repository. There
needs to be a way to control the visibility of prototypes as they
are developed and tested.
\item Coexistence in the current data ecosystem. The isolation of
today's exploratory visualization and data analysis environments is
a real obstacle to wider adoption of novel data analysis methods.
\item Modern interactive technologies. Current isolation notwithstanding,
there is demand for interactive visualization tools. In some respects,
the prevalence of data journalism has raised expectations very high,
but handmade, one-off designs are too expensive to most teams to support.
The main challenge is how to integrate these technologies so data
scientists and statisticians can program in them, \emph{and} business-facing
workers can operate them.
%% Scalable technologies like crossfilter exist. Want to support those in the context of an R environment. Need to support full 2-way communication between computing in the cloud and interaction on the web.
%% \item Other requirements e.g. Kandel's user study. Not our contribution but things we must address, mentioned in related work section. Heer/Agrawala design considerations for collaborative visual analytics. Commenting, annotation, modularity, granulatiry.
\end{itemize}

% What is the scope of our solution?
Our investigation focuses on the R language \cite{RCoreTeam:2013:R}
for statistical computing and graphics. R is one of the most popular
data analysis systems and is well-supported by a broad ecosystem
of packages, tools and other resources.
%
There are many existing R packages, including packages for high performance
computing in the cloud, domain specific data analysis, statistical computing,
and web graphics (though the latter has to be adapted for our architecture).
An additional requirement for our prototype is to work well with the rest
of this ecosystem.
